---
title: "EMTscoreData Package"
author: "haimei wen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EMTscoreData Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# 1. Introduction of EMTscoreData
EMTscoreData is an R ExperimentHub package that provides curated gene expression datasets for benchmarking and developing epithelial–mesenchymal transition (EMT) scoring methods. This package is prepared as the companion dataset for the upcoming EMTscore R package.

# 2. Installation
## Install the development version from GitHub
```{r install_github, eval=FALSE}
# install.packages("devtools")
#devtools::install_github("wenmm/EMTscore")
#devtools::install_github("wenmm/EMTscoreData")
```

## Install from Bioconductor (recommended once available):
```{r install}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("EMTscore")
#BiocManager::install("EMTscoreData")
```

# 3. Download data from the EMTscoreData R package
You can obtain the datasets included in this package through the `ExperimentHub` framework. To start, use query() to inspect all records associated with the `EMTscoreData` package, which will show you the available resources along with their corresponding identifiers. After identifying the items of interest, you can download them directly from the hub.
```{r}
library(EMTscoreData)
library(EMTscore)
library(ExperimentHub)
library(SummarizedExperiment)

eh = ExperimentHub()
query(eh , 'EMTscoreData')
```

## Download data using the unique identifier:
```{r}
A549_TNF <- eh[['EH10291']]
A549_EGF <- eh[['EH10292']]
A549_TGFB1 <- eh[['EH10293']]
```

# 4. Data analysis and visualization

## 4.1 Load R package and data
```{r}
library("EMTscore")
data("Panchy_et_al_E_signature", package = "EMTscore")
data("Panchy_et_al_M_signature", package = "EMTscore")
```

## 4.2 Prepare gene sets and save as GMT
```{r}
gene_sets <- list(
Panchy_et_al_E_signature = Panchy_et_al_E_signature$GeneName,
Panchy_et_al_M_signature = Panchy_et_al_M_signature$GeneName
)
write_gmt(gene_sets, "EM_signature.gmt")
```

## 4.3 Calculating EMT Scores for Datasets Stored in EMTscoreData
Calculated EMT scores can then be plotted against pseudotime to examine dynamic changes in epithelial–mesenchymal states.
```{r}
objects <- list(
A549_TGFB1 = A549_TGFB1,
A549_EGF   = A549_EGF,
A549_TNF   = A549_TNF
)
gmt <- system.file("extdata", "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION.v2025.1.Hs.gmt", package = "EMTscore")

seurat_objs <- add_EMT_score(objects, gmt_file = gmt, emt_name = "EMT_score", method = "AUCell",nnPCA_dim = 1)
p_AUCell <- plot_EMT_from_objects(seurat_objs, col_name = "Pseudotime", emt_score_col = "EMT_score")
p_AUCell
```

Calculate EMT scores with multiple EMT gene sets
```{r}
emt_name = c("Escore", "Mscore")
result <- add_EMT_score_multiple(objects, gmt_file = "EM_signature.gmt", emt_name, method = "nnPCA", nnPCA_dim = 1, cores = 10)

```


```{r}
sessionInfo()
```